Outliers Detection and Comparison of Origin-Destination Flows with Data Depth

### 2018-10-09 giscience_v2/giscience/giscience2018.tex
xx

\subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{Dummy keyword -- please provide 1--5 keywords}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Abstract

@@

\section{Introduction}
With the rapid rise in ubiquity of geolocation-aware sensors, knowledge discovery is greatly enhanced by extracting and mining interesting patterns from spatiotemporal Big Data in various domains.
In particular, the location-acquisition technologies generate large volumes of movement data, which were used to track people, animals, vehicles, and even natural phenomena.
Such data help us better model moving objects and reveal hidden patterns that are important to urban planning and its applications in characterizing urban human mobility \cite{jiang2012discovering}, accessibility of the activity space \cite{kwan1998space}, and sustainability of the urban systems \cite{alberti2003integrating}. 
Importantly, trajectory mining leads to solutions for addressing important research problems in different fields, such as urban planning \cite{mazimpaka15AGILE}, transportation \cite{chen13Percom}, environment \cite{devarakonda13SIGKDD}, and public security and safety \cite{buchin14JOSIS}.
This paper presents a new algorithm which not only estimates origination-destination (OD)'s flows anomaly but also conducts hypothesis testing between two sets of different OD flows. The algorithm is applied on the New York City taxi data set, where each record contains the origin and destination of each trip (without intermediate locations of the actual routes).
We believe the our methods for analyzing taxi trip data has a promising application in understanding crowd patterns and planning urban and transportation investments by the administrative authorities.

In recent years, researchers have investigated a variety of approaches to trajectory data mining.
In general, the contemporary trajectory mining methods can be classified into four categories: clustering, classification, frequent/group pattern mining, and outlier detection \cite{mazimpaka16JOSIS,zheng15ACMTIST}. These techniques can be used independently or combinatorially for trajectory mining applications. For this study, we focus on outlier detection of OD flows. Outlier detection aims to find trajectories that do not follow the typical flow of trajectory data sets for characterizing the connectivity between regions \cite{mazimpaka16JOSIS}. \cite{fontes13GeoInfo,lee08ICDE} use Euclidean distance to find outlier patterns from trajectories. \cite{pan13ACMGIS,liu12IJGIS} raise the questions of Euclidean distance approach due to the loss of local features and unavailability when external factors affect the trajectories (e.g., topography, land cover or weather condition).  \cite{pan13ACMGIS,liu12IJGIS} addresses this by using robust distance measurements (i.e., Mahalanobis distance \cite{pan13ACMGIS} and relative distance \cite{liu12IJGIS}). Further, structural features \cite{yuan11JCIS} and a data-induced random tree  \cite{zhang11UC} are exploited to detect anomalous trajectories instead of using distance or density. In a similar fashion, this study considers the center regions of OD flows with data depth in order to robustly detect the trajectory outliers.

In addition, utilizing visual analytics is a common approach to analyze OD flow data.
Visual representation of massive movement data enables comprehensive exploration of data and results in understanding complex flow trends.
Aggregation and generalization of movement data are frequently utilized \cite{andrienko08VAST,adrienko11IEEETVCG,guo14IEEETVCG}.
While visual analytics can help to extract inherent patterns from massive data, it is difficult to quantatively compare two sets of different OD flows based on a hypothesis testing.
In other words, it is complicated to comprehend how two OD flows differ and, more importantly, by how much.
In this light, this paper uses bivariate hypothesis testing methods based on data depth to understand how different two OD flow data sets are in terms of the amount of spatial extent and a shift in location.

It is worthy noting that flow mapping approaches frequently suffer from the modifiable areal unit problem (MAUP). For instance, it is not guaranteed that different aggregations via location can present coherent patterns.
Kernel-based flow estimation and smoothing are used to overcome different spatial resolution \cite{guo14IEEETVCG}. 
Instead of focusing on finding the best areal unit to partioning the urban space and aggregating the OD flows, this study chooses traffic analysis zones in New York City as a base unit. However, the mothods can be adapted to other choices of areal units.
In addition, New York City taxi trip data include the origin and destination within traffic analysis zones, which ignores the intermediate locations of the actual routes.
It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).

In summary, this paper presents a new algorithm which conducts outlier detection as well as a hypothesis testing from OD flows data. Our approach investigates the central regions of OD flows based on data depth to detect OD flows anomaly and conduct a hypothesis testing between two different OD flow data sets. The remainder of this paper is organized as follows: Section \ref{sec:methods} overviews how to detect OD flows outliers and conduct  a hypothesis testing between two different OD flows with the concept of data depth. 
Experimental design and the evaluation of the algorithm are presented in Section \ref{sec:experiments}. 
These results are discussed in Section \ref{sec:discussion}. 
Section 5 concludes with a summary and future perspectives.

%However these approaches take into account whole trajectories or sections of trajectories based on applications. This study focuses on OD flow data, which ignores the actual trajectory route. Additionally, it is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}). Further, this study considers spatial association of OD flow patterns with data depth to detect trajectory outliers. However, this study focuses on OD flow data, which ignores the actual trajectory route. The methods mentioned above consider the whole trajectory route. Further, New York City Taxi data have obivious  
%Clustering techniques are used to analyze the moving objects such as daily travel in a city and cyclone tracks \cite{rinzivillo12KI,gaffney07CD}. Human mobility is represented as a network with clustering technique \cite{liu09CUPUM}.

%Trajectory pattern mining.//
%clustering//
%Outliers//
%
%
%Trajectory clustering can be applied on either whole trajectories or sections of trajectories
%depending on the goal of analysis and the similarity function applied. For example, if
%the similarity function is defined as similar origin and similar destination, the trajectories
%can be clustered as a whole because the route followed is not important.
%
%
%visualization//
%
%Statistical testing//
%
%MAUP problems, uncertainty//
%
%make no attempt to 
%
%
%The  rapid growth of geospatial data results in enhancing knowledge discovery power by extracting interesting patterns from big data in various domains. There are many mining and analysis methodologies to extract new insights from observed data. In particular, this paper investigates spatio-temporal pattern maps that enable hypothesis testing to understand how groups differ and by how much. 
%
%In addition, our approach provides a method to comparing two independent groups based on data depth. The typical KDE can not provide its statistically significant difference between two different density maps. This study provides how different two data sets are in terms of the amount of scale and a shift in location. These features provide an interesting and useful perspective when trying to characterize how group differ. 
%However, the current study is specifically designed to deal with the entire area between two data sets. It is necessary to expand our approach to investigate local variations. Further, work might also test and compare with alternatives such as a spatial scan. 
%
%Our approach can take into account the overall structure of the data. This method can contribute to gaining a better understanding of data in a variety of applications such as influenza spread map or radiation level changes.
%
%The rapid growth of geospatial data can result in enhancing knowledge discovery power by
%extracting interesting patterns from big data in various domains. The identification of statistically significant hot spots has become a key problem that numerous businesses and organizations are attempting to solve. There are many spatial data mining and analysis methodologies to extract unusual patterns of data. This paper proposes a new spatiotemporal hypothesis testing using data depth. This study investigates how much two independent data sets are different in terms of the amount of spatial extent and a shift in location. These features provide an interesting and useful perspective when trying to characterize how groups differ. The experimental comparison with the leading state-of-the-art alternatives demonstrates that the proposed algorithm can not only take into account the overall structure of the data such as hot spots of drop-off locations in the New York City Taxi data, but also conducts statistical hypothesis testing. This method can contribute to gaining a better understanding of data in a variety of applications.

@@

\subsection{Data Depth}
Data depth measures the centrality of a point with regard to a given data sets in $\mathbb{R}^d$.
The notion of data depth (i.e., Tukey's halfspace depth) is originally developed by \cite{tukey75ICM}, which generalizes the univariate concept of ranking to multivariate data.
It reveals how deeply a point is located within a given data set by ordering their degree of centrality. 
In general, the halfspace depth (HD) of a point $x$ in  $\mathbb{R}^d$ is defined as the minimum probability, $P$ on  $\mathbb{R}^d$, associated with any closed halfspace containing $x$ \cite{liu00AS}. 
\begin{equation*}\label{eq:hd}
HD(x;P) = inf\{P(H): \text{H a closed halfspace}, x \in H\}\text{, } x \in \mathbb{R}^d.
\end{equation*}

@@

The property of halfspace depth is a center-outward ordering of points in  $\mathbb{R}^d$ and affine invariant \cite{Mosler13book}. These features serve as a useful tool in nonparametric inference, which lead to various applications such as data classification and clustering analysis
\cite{lange14fSP,jeong16acmgis}. There are a couple of approaches to calculate data depth: halfspace detpth \cite{rousseeuw96RSS}, projection depth \cite{wilcox03CSSC}, and simplicial depth \cite{liu90AS}. While the computational complexity of the projection approach is $\mathcal{O}(n^2)$ (where $n$ is the number of points), the computational complexity of simplicial depth is $\mathcal{O}(n^3)$. This can significantly increase execution time when $n$ is large. Thus, this paper uses the halfspace depth fuction proposed by \cite{rousseeuw96RSS} of which computation complexity is $\mathcal{O}(n\log{}n)$.

\subsection{OD Flow Outlier Detection Based on Depth}
The center-outward ordering in data depth is closely related to the detection of outliers.
The upper level sets of data depth in $\mathbb{R}^2$ form the central regions.
The most central region can be regarded as a median.
Conversely, the lower level sets of data depth, coincide with large distance from the center, can be regarded as outlyingness. \cite{rousseeuw99AS,aplpackR} utilizes this concept to generate a bag plot, which is analogous to the one-dimensional box plot based on data depth.
This paper uses the bag plot in order to identify the outliers of OD flows. 
Before explaining the method of outliers detection, we first introduce a couple of definitions.

@@

Based on these basic definitions, we plot the OD flows of New York City Taxi data (May 21 and July 01 2014) using a bag plot in Figure \ref{fig:bagplot_0521_0701}. In Figure \ref{fig:OD_0521}, the deepest depth of OD flows (i.e., depth median) is represented as a star symbol. This point is surrounded by a bag, which contains the half of OD flows (dark blue area). Magnifying the bag by a factor of 3 relative to depth median constructs a fence (light blue area). The fence is comparable with whiskers in an one-dimensional boxplot. The OD flows outside the fence are outliers (red color circles). The x-axis indicates the counts of forward OD flows and the y-axis indicates  the counts of reverse OD flows in Figure \ref{fig:OD_0521}. We can similarly present the bag plot of July 01 2014 in Figure \ref{fig:OD_0721}.

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/OD_0521.eps}
		\caption{May 21 2014}
		\label{fig:OD_0521}
	\end{subfigure}
	\hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 	%(or a blank line to force the subfigure onto a new line)
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/OD_0701.eps}
		\caption{July 01 2014}
		\label{fig:OD_0721}
	\end{subfigure}
    \hfill
	\begin{subfigure}[b]{0.49\textwidth}
	\includegraphics[width=\textwidth]{images/OD_0701_0521.eps}
	\caption{Combination of May 21 and July 01}
	\label{fig:OD_0721_0521}
	\end{subfigure}
    \hfill
	\begin{subfigure}[b]{0.49\textwidth}
	\includegraphics[width=\textwidth]{images/Outliers_high_0701_0521.eps}
	\caption{Detection of outliers for high volume of trips}
	\label{fig:OD_outliers_high}
    \end{subfigure}
	\begin{subfigure}[b]{0.53\textwidth}
	\includegraphics[width=\textwidth]{images/Outliers_rare_0701_0521.eps}
	\caption{Detection of outliers for atypical OD flow patterns}
	\label{fig:OD_outliers_rare}
    \end{subfigure}
	\caption{Outliers detection of OD flows using a bag plot}\label{fig:bagplot_0521_0701}	
\end{figure}

The bag plot is able to present the location (the depth median), spread (the size of bag), correlation (the orientation of the bag), and skewness (the shape of the bag and the fence) of data \cite{rousseeuw99AS}. For example, we can see which forward OD flows have higher counts compared to the reverse OD flows; the relatively linear correlation between forward OD flows and reverse OD flows; and the skewness of forward (reverse) OD flows in Figure \ref{fig:OD_0521}.

@@

\subsection{OD Flow Comparisons Based on Depth}
Data depth can compare bivariate data related with two independent groups \cite{liu93JASA}. 
A t-test can be used to compare means from two independent groups.
For example, the t-test reveals whether two OD flows' means are different at two different temporal ranges.
However, it is worth examining how groups differ in terms of scale.
The comparisons of central regions in data depth compare the marginal distribution, thereby considering the overall structure of the data \cite{wilcox03MBR}.

@@

For ease of understanding, Figure \ref{fig:com_mar_0329} presents the central regions of two OD flows.
One data set is OD flows for Saturday, March 29 2014 and the other is Saturday data sets which include March  1, 8, 15, 22, and April 5.
March 29 has the highest number of taxi trips of the year (i.e., 552,064 taxi trips).
The data set for the list of Saturdays has 2,621,703 taxi trips. The bootstrap method reveals that the confidence interval is 0.02467486 and 0.0595938. This confidence interval does not include zero, which results in rejecting the $H_0$. 
This means that the amount of scale is significantly changed between two OD flows.

In addition, it is apparent that the bootstrap method is a time consuming process.
We generate $2,000$ bootstrap samples. In order to scale up the bootstrap computation, we spread work across multiple nodes and cores by implementing an embarrassingly parallel R code.  Thus, we investigate resource pressures between serial and parallel approaches in the following section.

@@

\subsection{Data}
This study uses New York City Taxi data in 2014 to evaluate the effectiveness of the proposed approach.
Figure \ref{fig:taxizone} presents taxi analysis zones in New York City which indicates the origin and the destination IDs of OD flows. Figure \ref{fig:ODflows} shows OD flows on July 01.

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/taxizone.eps}
		\caption{2,250 taxi analysis zone in New York City}
		\label{fig:taxizone}
	\end{subfigure}
	\hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 	%(or a blank line to force the subfigure onto a new line)
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/July1st_pic2.png}
		\caption{OD flows on July 01 2014}
		\label{fig:ODflows}
	\end{subfigure}
	\caption{Experimental data: New York City Taxi data}\label{fig:taxi_data}	
\end{figure}

As a case study, this study used OD trips for the comparison between weekdays and weekends in June. The weekdays data set includes taxi trajectories on June 3, 10, 17, and 24. There are 1,721,655 taxi trips. The weekends data set includes taxi trajectories on June 8, 15, 22, and 29.  Its taxi trips are 1,593,480.

\subsection{Procedure}

The performance of the proposed method was compared with alternative methods. Trajectories anomalies detection, based on Mahalanobis distance \cite{pan13ACMSigspatial},  was used to compare the performance of outliers detection. The Mahalanobis distance considers the correlations of the data (e.g., two OD flows), which distinguishes it from Euclidean distance. According to \cite{pan13ACMSigspatial}, the anomaly detection threshold can be defined as below.

@@

With regard to data cleaning process, this study used Hadoop with Pig. R was used to implement the code in Amazon Web Service  and Bridges supercomputer resources. Further, we used OD flows that have more than 10 trips. It is reasonable to remove the majority of low number of OD flows, which may distort the view of the data.

@@

\subsection{Case study: weekdays vs weekends}
\subsubsection{Outlier Detection}
The bag plot presented OD flow outliers on weekdays and weekends separately in Figure \ref{fig:week_weekends_bag}. The outliers are detected by considering forward OD flows and reverse OD flow together.

@@

In order to find the difference between two data sets, we considered two forward OD flows together with the bag plot. Then we found the outliers of OD flows in Figure \ref{fig:weekdays_high}. In particular, the outliers with rectangle symbols indicates that these OD flows have high volume of taxi trips during weekdays and weekends. These outliers are superimposed on a map with red lines in Figure \ref{fig:weekdays_high_map}. The yellow lines presents OD flows except the high volume of OD flows on weekends.

@@

\subsection{Scalability}

>>>>

\subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{OD Analysis, Trajectory Data Mining, Data Depth, Outliers Detection}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
The advances in location-aware technologies have generated a huge volume of trajectory data. In particular, Origin-Destination (OD) trajectories provide rich information to understand urban flow and transport demand. This study presents a new methodology to detect OD flows outliers and conduct hypothesis testing between two OD flows in terms of the variations of spatial extent (i.e., spread). The proposed method is based on data depth, which measures the centrality and outlyingness of a point with respect to a given data set in $\mathbb{R}^d$. Based on the center-outward ordering property, it is possible to analyze the underlying OD flows characteristics such as location, outlyingness, and  spread. The proposed method is compared with Mahalanobis distance approach to detect OD anomalies, and F-test to identify the difference in scale. Empirical evaluation has demonstrated that the proposed method can identify OD flows outliers in an interactive way. Further, it can provide new perspectives by considering the overall structure of data when comparing two different OD flows in scale.

@@

\section{Introduction}
With the rapid rise in ubiquity of geolocation-aware sensors, knowledge discovery is greatly enhanced by extracting and mining interesting patterns from spatiotemporal big data in various domains.
In particular, the location-acquisition technologies generate large volumes of movement data, which are used to track people, animals, vehicles, and even natural phenomena.
Such data help us better model moving objects and reveal hidden patterns that are important to urban planning and its applications in characterizing urban human mobility \cite{yin2017depicting}, accessibility of the activity space \cite{kwan1998space}, and sustainability of the urban systems \cite{alberti2003integrating}.
Importantly, trajectory mining leads to solutions for addressing important research problems in different fields, such as urban planning \cite{mazimpaka15AGILE}, transportation \cite{chen13Percom}, environment \cite{devarakonda13SIGKDD}, and public security and safety \cite{buchin14JOSIS}.

This paper presents a new algorithm which not only estimates origination-destination (OD)'s flows anomaly but also conducts hypothesis testing between two sets of different OD flows. The algorithm is applied on the New York City taxi data set, where each record contains the origin and destination of each trip (without intermediate locations of the actual routes).
We believe that our methods for analyzing taxi trip data have a promising application in understanding crowd patterns, and planning urban and transportation investments by the administrative authorities.

In recent years, researchers have investigated a variety of approaches to trajectory data mining.
In general, the contemporary trajectory mining methods can be classified into four categories: clustering, classification, frequent/group pattern mining, and outlier detection \cite{mazimpaka16JOSIS,zheng15ACMTIST}. These techniques can be used independently or combinatorially for trajectory mining applications. For this study, we focus on outlier detection of OD flows. Outlier detection aims to find trajectories that do not follow the typical flow of trajectory data sets for characterizing the connectivity between regions \cite{mazimpaka16JOSIS}. \cite{fontes13GeoInfo,lee08ICDE} use Euclidean distance to find outlier patterns from trajectories. \cite{pan13ACMGIS,liu12IJGIS} raise the questions of Euclidean distance approach due to the loss of local features and unavailability when external factors affect the trajectories (e.g., topography, land cover or weather condition). \cite{pan13ACMGIS,liu12IJGIS} addresses this issue by using robust distance measurements (i.e., Mahalanobis distance \cite{pan13ACMGIS} and relative distance \cite{liu12IJGIS}). Further, structural features \cite{yuan11JCIS} and a data-induced random tree \cite{zhang11UC} are exploited to detect anomalous trajectories instead of using distance or density. In a similar fashion, this study considers the center regions of OD flows with data depth in order to robustly detect the OD flow outliers.

In addition, utilizing visual analytics is a common approach to analyze OD flow data.
Visual representation of massive movement data enables comprehensive exploration of data and results in understanding complex flow trends.
Aggregation and generalization of movement data are frequently utilized \cite{andrienko08VAST,adrienko11IEEETVCG,guo14IEEETVCG}.
While visual analytics can help extract inherent patterns from massive data, it is difficult to quantitatively compare two sets of different OD flows based on a hypothesis testing.
In other words, it is complicated to comprehend how two OD flows differ and, more importantly, by how much.
In this light, this paper uses bivariate hypothesis testing methods based on data depth to understand how different two OD flow data sets are in terms of the amount of spatial extent.

It is worth noting that flow mapping approaches frequently suffer from the modifiable areal unit problem (MAUP). For instance, it is not guaranteed that different aggregations via location can present coherent patterns.
Kernel-based flow estimation and smoothing are used to overcome different spatial resolution \cite{guo14IEEETVCG}. 
Instead of focusing on finding the best areal unit to partitioning the urban space and aggregating the OD flows, this study chooses traffic analysis zones in New York City as a base unit. However, the method can be adapted to other choices of areal units.
In this study, New York City taxi trip data include the origin and destination within traffic analysis zones, which ignores the intermediate locations of the actual routes.
It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).

In summary, this paper presents a new algorithm which conducts outlier detection as well as a hypothesis testing from OD flows data. Our approach investigates the central regions of OD flows, based on data depth, to detect OD flows anomaly and conduct a hypothesis testing between two different OD flow data sets. The remainder of this paper is organized as follows: Section \ref{sec:methods} overviews how to detect OD flows outliers and conduct  a hypothesis testing between two different OD flows with the concept of data depth. 
Experimental design and the evaluation of the methods are presented in Section \ref{sec:experiments}. 
These results are discussed in Section \ref{sec:discussion}. 
Section 5 concludes with a summary and future perspectives.

%With the rapid rise in ubiquity of geo-tagged sensors, knowledge discovery power is enhanced by extracting interesting patterns from spatiotemporal big data in various domains. In particular, the location-acquisition technologies generate large volumes of movement data such as people, animals, vehicles and natural phenomena. These big movement data help us understand moving objects, and discover the hidden patterns. Thus, trajectory mining leads to solutions for important research problems in different fields such as urban planning \cite{mazimpaka15AGILE}, transportation \cite{chen13Percom}, environment \cite{devarakonda13SIGKDD}, and public security and safety \cite{buchin14JOSIS}.
%
%This paper presents a new algorithm which not only estimates origination-destination (OD)'s flows anomaly but also conducts hypothesis testing between two different OD flows. This paper utilizes a New York City Taxi data set that has the origin and destination of each movement without the whole actual trajectory routes. The analysis of digital footprints obtained from taxi data has a crucial role in  understanding crowd patterns and planning urban and transportation investments in administrative authorities.

%In recent years, researchers have investigated a variety of approaches to trajectory data mining. There are generally four categories that classify trajectory mining methods: clustering, classification, frequent/group pattern mining, and outlier detection \cite{mazimpaka16JOSIS,zheng15ACMTIST}. These techniques can be used independently or combinatorially for trajectory mining application problems. 
%
%In particular, this study focuses on outlier detection of OD flows. Outlier detection aims to find trajectories that do not follow the typical flow of trajectory data sets for characterizing the connectivity between regions \cite{mazimpaka16JOSIS}. \cite{fontes13GeoInfo,lee08ICDE} use Euclidean distance to find outlier patterns from trajectories.  \cite{pan13ACMGIS,liu12IJGIS} raise the questions of Euclidean distance approach due to the loss of local features and unavailability when external factors affect the trajectories (e.g., topography, land cover or weather condition).  \cite{pan13ACMGIS,liu12IJGIS} addresses this by using robust distance measurements (i.e., Mahalanobis distance \cite{pan13ACMGIS} and relative distance  \cite{liu12IJGIS}). Further, structural features  \cite{yuan11JCIS} and a data-induced random tree  \cite{zhang11UC} are exploited to detect anomalous trajectories instead of using distance or density. In a similar fashion, this study considers the center regions of OD flows with data depth in order to robustly detect the trajectory outliers. 
%
%
%
%In addition, visual analytics are a common approach to analyze OD flow data. Visual representation of massive movement data enables comprehensive exploration of data and results in understanding complex flow trends. Aggregation and generalization of movement data are frequently utilized \cite{andrienko08VAST,adrienko11IEEETVCG,guo14IEEETVCG}. While visual analytics can help to extract inherent patterns from massive data, it is difficult to compare two different OD flows based on a hypothesis testing. In other words, it is complicated to comprehend how two OD flows differ and by how much. This paper uses bivariate hypothesis testing methods based on data depth to understand how different two OD flow data sets are in terms of the amount of spatial extent. % and a shift in location. 
%
%
%Further, flow mapping approaches frequently suffer from the modifiable area unit problem (MAUP). For instance, it is not guaranteed that different aggregations via location can present coherent patterns. Kernel-based flow estimation and smoothing are used to overcome different spatial resolution \cite{guo14IEEETVCG}. This study uses traffic analysis zones in New York City as a base unit. Thus, it does not consider MAUP problem. In addition, New York City Taxi data includes the origin and destination within traffic analysis zones, which ignores the actual trajectory route. It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).
%
%In summary, this paper presents a new algorithm which conducts outlier detection as well as a hypothesis testing from OD flows data. Our approach investigates the central regions of  OD flows based on data depth to detect OD flows anomaly and conduct a hypothesis testing between two different OD flow data sets. 
%
%In the remainder of this paper, Section \ref{sec:methods} overviews how to detect OD flows outliers and conduct  a hypothesis testing between two different OD flows with the concept of data depth. Experimental design and the evaluation of the proposed method are presented in Section \ref{sec:experiments}. These results are discussed in Section \ref{sec:discussion}. Section \ref{sec:conclusions} concludes with a summary and future perspectives.

@@

\subsection{Data Depth}
Data depth measures the centrality of a point with regard to a given data sets in $\mathbb{R}^d$. The notion of data depth (i.e., Tukey's halfspace depth) is originally developed by \cite{tukey75ICM}, which generalizes the univariate concept of ranking to multivariate data. It presents how deeply a point is located within a given data set by ordering their degree of centrality.

In general, the halfspace depth (HD) of a point $x$ in $\mathbb{R}^d$ is defined as the minimum probability, $P$ on $\mathbb{R}^d$, associated with any closed halfspace containing $x$ \cite{liu00AS}. 
\begin{equation*}\label{eq:hd}
HD(x;P) = inf\{P(H): \text{H is a closed halfspace}, x \in H\}\text{, } x \in \mathbb{R}^d.
\end{equation*}

@@

The property of halfspace depth is a center-outward ordering of points in  $\mathbb{R}^d$ and affine invariant \cite{Mosler13book}. These features serve as a useful tool in nonparametric inference, which lead to various applications such as data classification and clustering analysis
\cite{lange14fSP,jeong16acmgis}. There are a couple of approaches to calculate data depth: halfspace depth \cite{rousseeuw96RSS}, projection depth \cite{wilcox03CSSC}, and simplicial depth \cite{liu90AS}. While the computational complexity of the projection approach is $\mathcal{O}(n^2)$ (where $n$ is the number of points), the computational complexity of simplicial depth is $\mathcal{O}(n^3)$. This can significantly increase execution time when $n$ is large. Thus, this paper uses the halfspace depth function proposed by \cite{rousseeuw96RSS} of which computation complexity is $\mathcal{O}(n\log{}n)$.

\subsection{OD Trajectory Outlier Detection Based on Depth}
The center-outward ordering in data depth is closely related to the detection of outliers. The upper level sets of data depth in $\mathbb{R}^2$ form the central regions. The most central region can be regarded as a median. Conversely, the lower level sets of data depth, coincide with large distance from the center, can be regarded as outlyingness. \cite{rousseeuw99AS,aplpackR} utilizes this concept to generate a bag plot, which is analogous to the one-dimensional box plot based on data depth. This paper uses the bag plot in order to identify the outliers of OD flows. Before explaining the method of outliers detection, we first introduce a couple of definitions.

@@

Based on these basic definitions, we plot the OD flows of New York City Taxi data (May 21 and July 1 2014) using a bag plot in Figure \ref{fig:bagplot_0521_0701}. In Figure \ref{fig:OD_0521}, the deepest depth of OD flows (i.e., depth median) is represented as a star symbol. This point is surrounded by a bag, which contains the half of OD flows (dark blue area). Magnifying the bag by a factor of 3 relative to depth median constructs a fence (light blue area). The fence is comparable with whiskers in an one-dimensional boxplot. The OD flows outside the fence are outliers (red color circles). The x-axis indicates the counts of forward OD flows and the y-axis indicates  the counts of reverse OD flows in Figure \ref{fig:OD_0521}. We can similarly present the bag plot of July 1 2014 in Figure \ref{fig:OD_0721}.

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/OD_0521.eps}
		\caption{May 21 2014}
		\label{fig:OD_0521}
	\end{subfigure}
	\hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 	%(or a blank line to force the subfigure onto a new line)
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/OD_0701.eps}
		\caption{July 1 2014}
		\label{fig:OD_0721}
	\end{subfigure}
    \hfill
	\begin{subfigure}[b]{0.49\textwidth}
	\includegraphics[width=\textwidth]{images/OD_0701_0521.eps}
	\caption{Combination of May 21 and July 1}
	\label{fig:OD_0721_0521}
	\end{subfigure}
    \hfill
	\begin{subfigure}[b]{0.49\textwidth}
	\includegraphics[width=\textwidth]{images/Outliers_high_0701_0521.eps}
	\caption{Detection of outliers with high volume of trips}
	\label{fig:OD_outliers_high}
    \end{subfigure}
	\begin{subfigure}[b]{0.53\textwidth}
	\includegraphics[width=\textwidth]{images/Outliers_rare_0701_0521.eps}
	\caption{Relative outliers detection}
	\label{fig:OD_outliers_rare}
    \end{subfigure}
	\caption{Outliers detection of OD flows using a bag plot}\label{fig:bagplot_0521_0701}	
\end{figure}

The bag plot is able to present the location (the depth median), spread (the spatial extent of bag), correlation (the orientation of the bag), and skewness (the shape of the bag and the fence) of data \cite{rousseeuw99AS}. For example, we can see which forward OD flows have higher counts compared to the reverse OD flows; the relatively linear correlation between forward OD flows and reverse OD flows; and the skewness of forward (reverse) OD flows in Figure \ref{fig:OD_0521}.

@@

\subsection{OD Trajectory Comparisons Based on Depth}
Data depth can compare bivariate data related with two independent groups. A t-test can be used to compare  means from two independent groups. For example, the t-test reveals whether two OD flows' means  are different at two different temporal ranges. However, it is worth examining how groups differ in terms of scale (spread). The comparisons of central regions in data depth compare the marginal distribution, thereby considering the overall structure of the data \cite{wilcox03MBR}.

@@

For ease of understanding, Figure \ref{fig:com_mar_0329} presents the central regions of two OD flows. One data set is OD flows for Saturday, March 29 2014 and the other is  the data set of Saturdays which include March  1, 8, 15, 22, and April 5. March 29 has the highest number of taxi trips of the year (i.e., 552,064 taxi trips).  The data set for the list of Saturdays has 2,621,703 taxi trips. The bootstrap method reveals that the confidence interval is 0.02467486 and 0.0595938. This confidence interval does not include zero, which results in rejecting the $H_0$. This means that the amount of scale is significantly changed between two OD flows. Further, it is possible to see the OD flows from a list of Saturdays is nested within the OD flows corresponding to March 29. This is additional perspective based on data depth comparisons.

In addition, it is apparent that the bootstrap method is a time consuming process. We generate $2,000$ bootstrap samples. In order to scale up the bootstrap computation, we distribute work across multiple nodes and cores by implementing an embarrassingly parallel R code.  Thus, we can effectively improve execution efficiency.
% investigate resource pressures between serial and parallel approaches in the following section.

@@

\subsection{Data}
This study uses New York City Taxi data in 2014 to evaluate the effectiveness of the proposed approach.
Figure \ref{fig:taxizone} presents taxi analysis zones in New York City which indicate the origin and the destination IDs of OD flows. Figure \ref{fig:ODflows} shows OD flows on July 1. Red lines indicate the dominated OD flows.

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/taxizone.eps}
		\caption{2,250 taxi analysis zones in New York City}
		\label{fig:taxizone}
	\end{subfigure}
	\hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 	%(or a blank line to force the subfigure onto a new line)
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/July1st_pic2.png}
		\caption{OD flows on July 1 2014}
		\label{fig:ODflows}
	\end{subfigure}
	\caption{Experimental data: New York City Taxi data}\label{fig:taxi_data}	
\end{figure}

As a case study, this study used OD flows during weekdays and weekends in June, 2014. The weekdays data set includes taxi trajectories on June 3, 10, 17, and 24. There are 1,721,655 taxi trips. The weekends data set includes taxi trajectories on June 8, 15, 22, and 29.  Its taxi trips are 1,593,480.

\subsection{Procedure}
The performance of the proposed method was compared with alternative methods. Trajectories anomalies detection, based on Mahalanobis distance \cite{pan13ACMGIS},  was used to compare the performance of outliers detection. The Mahalanobis distance considers the correlations of the data (e.g., two OD flows), which distinguishes it from Euclidean distance. According to \cite{pan13ACMGIS}, the anomaly detection threshold can be defined as below.

@@

With regard to data cleaning process, this study used Hadoop with Pig. We have developed a Hadoop program to handle the large volume of data (~173 million taxi trip records), remove trips with invalid OD coordinates, and assign each OD locations into the corresponding traffic analysis zone.  
R was used to implement the OD flow outlier detection. The computing environment uses Amazon Web Service and Bridges supercomputer at the Pittsburgh Supercomputing Center. Further, we used OD flows that have more than 10 trips. It is reasonable to remove the majority of low number of OD flows, which may distort the view of the data. All the code will be released as open source (the link to the code will added later and currently is available upon request).

@@

\subsection{Case study: weekdays vs weekends}
\subsubsection{Outlier Detection}
The bag plot presented OD flow outliers on weekdays and weekends separately in Figure \ref{fig:week_weekends_bag}. The outliers are detected by considering forward OD flows and reverse OD flows together.

@@

In order to find the difference between two data sets, we considered two forward OD flows together with the bag plot. Then we found the outliers of OD flows in Figure \ref{fig:weekdays_high}. In particular, the outliers with rectangle symbols indicate that these OD flows have high volume of taxi trips during weekdays and weekends. These outliers are superimposed on a map with red lines in Figure \ref{fig:weekdays_high_map}. The yellow lines present OD flows except the high volume of OD flows on weekends.

@@

%\subsection{Scalability}
%
%\begin{figure}
%	\centering
%	\begin{subfigure}[b]{0.7\textwidth}
%		\includegraphics[width=\textwidth]{images/scalability.eps}
%	\end{subfigure}
%	\caption{Measure of time reduction. }\label{fig:scalability}	
%\end{figure}
